Parametricity for Primitive Nested Types and GADTs

### 2020-12-24 free-theorems-nesteds/20-mscs.tex
more work on Sec 8

If $\Gamma$ is a finite subset of\, $\tvars$, $\Phi$ is a finite
subset of\, $\fvars$, $\overline{\alpha}$ is a finite subset of\,
$\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
\Phi$, then the set $\mcF$ of {\em functorial expressions} over
$\Gamma$ and $\Phi$ are given in Definition~\ref{def:wftypes}. The
notation there entails that an application $F F_1...F_k$ is allowed
only when $F$ is a type variable of arity $k$, or $F$ is a
subexpression of the form $\mu \phi^{k}.\lambda
\alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity $k$ then $F$ must
be applied to exactly $k$ arguments.  Accordingly, an overbar
indicates a sequence of subexpressions whose length matches the arity
of the type applied to it.  Requiring that types are always in such
\emph{$\eta$-long normal form} avoids having to consider
$\beta$-conversion at the level of types. In a subexpression
$\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
the variables in $\ol{\alpha}$ in $F$ and $G$; intuitively,
$\Nat^{\ol{\alpha}}F\,G$ represents the type of a natural
transformation in $\ol{\alpha}$ from the functor $F$ to the functor
$G$.  In a subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$
operator binds all occurrences of the variable $\phi$, and the
$\lambda$ operator binds all occurrences of the variables in
$\ol{\alpha}$, in the body $F$.

@@

\begin{defi}\label{def:wftypes}
The formation rules for the set $\F$ of\, {\em well-formed types} are

@@

\begin{defi}\label{def:lke}
If $F : \set^k \to \set$ and $\ol{K} : \set^k \to \set^h$ are functors
over $\set$, then the {\em left Kan extension of $F$ along $\ol K$} is
a functor $\mathit{Lan}_{\ol K}\,F : \set^h \to \set$ together with a
natural transformation $\eta : F \to (\mathit{Lan}_{\ol K}\,F)
\circ \ol K$
%are initial among such pairs of functor and natural transformations,
%i.e.,
such that, for every functor $L : \set^h \to \set$ and natural
transformation $\epsilon : F \to L \circ \ol K$, there exists a unique
natural transformation $\mu : \mathit{Lan}_{\ol K}\,F \to L$ such that
$(\mu {\ol K}) \circ \eta = \epsilon$. This is depicted in the
following diagram:

@@

\[
\begin{tikzcd}[row sep = huge]
\set^k
\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
\ar[rd, "{\ol K}"']
&& \set \\
& \set^h
\ar[Rightarrow, bend right = 10, from=F, "{\eta}" description]
\ar[ur, bend left, "{\mathit{Lan}_K\,F}"{name=Lan, description}]
\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\epsilon}"]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
\end{tikzcd}
\]

@@

A third representation of left Kan extensions in locally presentable
categories is given in terms of colimits. Writing ${\mathcal C}_0$ for
the full subcategory of finitely presentable objects in the locally
presentable category $\mathcal C$, the left Kan extension can be
expressed as
\[(\mathit{Lan}_{\ol K} F)\ol A \;=\; \colim{(\ol{X : \,{\mathcal C}_0},\,
  \ol{f : K{\ol X} \to A})}{F \ol X}\]

Thus, if $F : \set^k \to \set$, $\ol{K} :
  \set^k \to \set^h$, $\ol{A} : \set^h$, and $\set_0$ is the full
  subcategory of finitely presentable objects in $\set$, --- i.e., is
  the category finite sets --- we  have that
\[(\mathit{Lan}_K F) \ol{A} = \colim{(\ol{S} : \set_0^k, \,\ol{f : K\ol{S} \to
    A})}{F \ol{S}}\]
If $\ol{S} : \set_0^k$ and $\ol{f : K\ol{S} \to A}$, let $j_{\ol{S},
  \ol{f}} : F \ol{S} \to (\mathit{Lan}_KF) \ol{A}$ be the morphism
indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit.
Then for any $h$-tuple of functions $\ol{g : A \to B}$, the functorial
action $(\mathit{Lan}_K F) \ol{g}$ is the unique function from
$(\mathit{Lan}_K F) \ol{A}$ to $(\mathit{Lan}_K F) \ol{B}$ such that,
for every $\ol{S} : \set^k_0$ and $\ol{f : K\ol{S} \to A}$,\, 
$(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
  \ol{g\, \circ f}}$ holds.

We can similarly represent left Kan extensions of functors over $\rel$
in terms of colimits. In that setting we will denote the morphism
mapping the cocone into the colimit $\iota$. We will make good use of
both of these representations in Section~\ref{sec:term-sem} below.

@@

\[
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
\DisplayProof
\]
and
\[
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
\DisplayProof
\]

@@

\noindent
is defined to be
\begin{multline*}
\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}\rho'} \\
\circ\;
  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}f}
\end{multline*}
or, equivalently by naturality,
\begin{multline*}
\big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}f} \\
\circ\;
  \big(\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}\,
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
\end{multline*}

@@

Recalling the notation for the colimit representation of left Kan
extensions from Section~\ref{sec:lke}, we define the set and
relational interpretations of the terms from the introduction and
elimination rules for $\Lan$ types from Section~\ref{sec:ext-calc} by 
{\color{red} CAUTION: semantics don't match the rules!}
\begin{multline*}
\setsem{\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol{K}}\, F)\, \ol{K[\ol{\alpha := A}]}}\rho\;d \\
= j_{\ol{\setsem{\Gamma;\Phi \vdash A}\rho},
  \ol{\id_{\setsem{\Gamma; \Phi \vdash K[\ol{\alpha := A}]}\rho}}}\,
(\setsem{\Gamma; \Phi~|~\Delta \vdash t : F[\ol{\alpha := A}]} \rho\, d)
\end{multline*}
and
\[
\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta 
: \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\; G}\rho
= \lambda \ol{N} \, \ol{B}.\, (\mu_{\ol{N}})_{\ol{B}}
\]
respectively, where
\[\mu_{\ol{N}} : \mathit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho
  {[\ol{\alpha := \_} ]}}} \setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho {[\Phi, \ol \alpha := \ol N, \_ ]} \to \setsem{\Gamma;
    \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \_]\]
  is the unique natural transformation such that
\begin{equation}\label{eq:defn-of-mu-sub-N}
    (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash \eta : \Nat^{\Phi,
    \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}} =
(\mu_{\ol{N}} \, \ol{\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := \_}]}) \circ \epsilon_{\ol{N}}
  \end{equation}
{\color{red} $\eta$ vs $\epsilon$}
as given by the universal property of the left Kan extension.
Here, $\epsilon_{\ol{N}}$ is the natural transformation associated
with the left Kan extension, as shown in the following diagram.
\[\begin{tikzcd} [column sep = large, row sep = large]
\set^{|\ol \alpha|}
\ar[rr, "\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol
      \alpha := \ol N, \_ ]}"{name=F}] 
\ar[dr, "\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}"{left}] && \set \\ 
&\set^{|\ol K|} \ar[ur, "\mathit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := \_} ]}}}\,
\setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha :=
    \ol N, \_ ]}"{right}]
\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
          from=F, to=2-2,  "\epsilon_{\ol{N}}"{}]
\end{tikzcd}\]
The relational interpretations are define analogously.

Having defined the syntax and both the set and relational semantics for $\Lan$ types and terms,
we need to check that these definitions yield a parametric model.
The first step would be to extend Lemma~\ref{lem:rel-transf-morph} to $\Lan$ types,
by proving a new clause of the inductive step of the proof.
Unfortunately, it turns out that, in general, this is not possible, as the following example shows.

\begin{prop}\label{prop:full-lan}
Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
Assume that $L : \set^k \to \set$ is a functor such that the following properties hold:
\begin{enumerate}
\item for all $\ol{A : \set}$, we have that $L \ol{A} \subseteq (\textit{Lan}_{\ol K}\, F)\ol{A}$
\item for all $\ol{f : A \to B}$ and $\ol{x \in A}$, we have that $(\textit{Lan}_{\ol K}\, F)\ol{f}|_{L \ol{A}}\, x \in L \ol{B}$
\item for all $\ol{A : \set}$ and $y \in F \ol{A}$, we have that $\epsilon_{\ol{A}} y \in L ( \ol{K \ol{A}} )$
\end{enumerate}
Then $L = \textit{Lan}_{\ol K}\, F$.
\end{prop}
\begin{proof}
For every $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K}\, F)\ol{A}$
there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$ and $w \in F \ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}} w$.
Thus, $z = \iota_{\ol{S}, \ol{f}} w = (\textit{Lan}_{\ol K}\, F) \ol{f} (\epsilon_{\ol{S}} w)$;
by the third assumption we have that $\epsilon_{\ol{S}} w \in L ( \ol{K \ol{S}} )$
and by the second assumption we have that $z \in L \ol{A}$.
We can than conclude that $L \ol{A} = (\textit{Lan}_{\ol K}\, F)\ol{A}$.
\end{proof}

\begin{exa}
We provide a counterexample to Lemma~\ref{lem:rel-transf-morph} for $\Lan$ types.
Consider the type $\emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha$.
Then, using the colimit formula for left Kan extensions, we have
\[
\begin{array}{rl}
\relsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } \rho [\alpha := (1, 0, 0)]
&= (\textit{Lan}_{\Eq_1}\, \Eq_1)(1, 0, 0) \\
&= \colim{f : \Eq_1 \to (1, 0, 0)}{\Eq_1} \\
&= (0, 0, 0)
\end{array}
\]
because there exists no morphism $f : \Eq_1 \to (1, 0, 0)$ in $\rel$.
But, considering the set interpretation on the first projection of the relation environment, we have
\[
\begin{array}{rl}
\setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } ( \pi_1 ( \rho [\alpha := (1, 0, 0)] ) )
&= \setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet}\, \onet) \alpha } (\pi_1 \rho) [\alpha := 1] \\
&= (\textit{Lan}_{1}\, 1) 1 \\
&= \colim{1 \to 1}{1} \\
&= 1
\end{array}
\]
because there exist a unique arrow $1 \to 1$ in $\set$, which is the identity on $1$.
\end{exa}

The issue lies in the fact that the functor along which we extend contains constants (such as $\onet$).
Indeed, if we assume that $\ol{\emptyset; \overline{\alpha} \vdash K}$ are polynomial functors not containing constants,
then we can prove that the relational interpretation of
$\Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$
is fibered over its set interpretation, as proven in the following proposition.

\begin{prop}
Assume that the types in $\ol{K}$ are polynomial and contain no constants
(i.e., they are formed only by the functorial type variables $\ol\alpha$, $+$ and $\times$).
Then
\[
\pi_i (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
)
=
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
\]
for every relation environment $\rho$ and $i = 1, 2$.
\end{prop}
\begin{proof}
Notice that, by the colimit formula,
\[
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
=
\colim{\ol{R}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } \rho[\ol{\alpha := R}]
}
\]
(where we call $\iota$ the cocone of the colimit)
and thus
\[
\pi_i (
\relsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } \rho
)
=
\colim{\ol{R : \rel_0}, \ol{m : \relsem{ \emptyset; \ol\alpha \vdash K } \rho[\ol{\alpha := R}] \to \relsem{ \Gamma; \Phi \vdash A } \rho}}{
\relsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := \pi_i R}]}
\]
(where the cocone of the colimit is $\pi_i \iota$),
whereas, again by the colimit formula,
\[
\setsem{ \Gamma; \Phi \vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A} } (\pi_i \rho)
=
\colim{\ol{S : \set_0}, \ol{f : \setsem{ \emptyset; \ol\alpha \vdash K } (\pi_i \rho)[\ol{\alpha := S}] \to \setsem{ \Gamma; \Phi \vdash A } (\pi_i \rho)}}{
\setsem{ \Gamma; \Phi, \ol\alpha \vdash F } (\pi_i \rho) [\ol{\alpha := S}]}
\]
(where we call $j$ the cocone of the colimit).

>>>>

If $\Gamma$ is a finite subset of\, $\tvars$, $\Phi$ is a finite
subset of\, $\fvars$, $\overline{\alpha}$ is a finite subset of\,
$\fvars^0$ disjoint from $\Phi$, and $\phi^k \in \fvars^k \setminus
\Phi$, then the set $\mcF$ of {\color{red} {\em functorial expressions}} over
$\Gamma$ and $\Phi$ are given in Definition~\ref{def:wftypes}. The
notation there entails that an application $F F_1...F_k$ is allowed
only when $F$ is a type variable of arity $k$, or $F$ is a
subexpression of the form $\mu \phi^{k}.\lambda
\alpha_1...\alpha_k.F'$. Moreover, if $F$ has arity $k$ then $F$ must
be applied to exactly $k$ arguments.  Accordingly, an overbar
indicates a sequence of subexpressions whose length matches the arity
of the type applied to it.  Requiring that types are always in such
\emph{$\eta$-long normal form} avoids having to consider
$\beta$-conversion at the level of types. In a subexpression
$\Nat^{\ol{\alpha}}F\,G$, the $\Nat$ operator binds all occurrences of
the variables in $\ol{\alpha}$ in $F$ and $G$; intuitively,
$\Nat^{\ol{\alpha}}F\,G$ represents the type of a natural
transformation in $\ol{\alpha}$ from the functor $F$ to the functor
$G$.  In a subexpression $\mu \phi^k.\lambda \ol{\alpha}.F$, the $\mu$
operator binds all occurrences of the variable $\phi$, and the
$\lambda$ operator binds all occurrences of the variables in
$\ol{\alpha}$, in the body $F$.

@@

\begin{defi}\label{def:wftypes}
The formation rules for the set $\F$ of\, {\color{red} {\em
    well-formed types}} are

@@

\begin{defi}\label{def:lke}
If $F : \set^k \to \set$ and $\ol{K} : \set^k \to \set^h$ are functors
over $\set$, then the {\em left Kan extension of $F$ along $\ol K$} is
a functor $\mathit{Lan}_{\ol K}\,F : \set^h \to \set$ together with a
natural transformation $\eta : F \to (\mathit{Lan}_{\ol K}\,F)
\circ \ol K$
such that, for every functor $G : \set^h \to \set$ and natural
transformation $\gamma : F \to G \circ \ol K$, there exists a unique
natural transformation $\mu : \mathit{Lan}_{\ol K}\,F \to G$ such that
$(\mu {\ol K}) \circ \eta = \gamma$. This is depicted in the
following diagram:

@@

\[
\begin{tikzcd}[row sep = huge]
\set^k
\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
\ar[rd, "{\ol K}"']
&& \set \\
& \set^h
\ar[Rightarrow, bend right = 10, from=F, "{\eta}" description]
\ar[ur, bend left, "{\mathit{Lan}_K\,F}"{name=Lan, description}]
\ar[ur, bend right, "{G}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\gamma}"]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
\end{tikzcd}
\]

@@

A third representation of left Kan extensions in locally presentable
categories is given in terms of colimits. Writing ${\mathcal C}_0$ for
the full subcategory of finitely presentable objects in the locally
presentable category $\mathcal C$, the left Kan extension can be
expressed as
\[
(\mathit{Lan}_{\ol K} F)\ol A \;=\; \colim{\ol{X : \,{\mathcal
        C}_0},\, \ol{f : K{\ol X} \to A}}{F \ol X}\]
Thus, if $F : \set^k \to \set$, $\ol{K} :
  \set^k \to \set^h$, $\ol{A} : \set^h$, and $\set_0$ is the full
  subcategory of finitely presentable objects in $\set$, --- i.e., is
  the category finite sets --- we  have that
\begin{equation}\label{eq:colim-form}
(\mathit{Lan}_K F) \ol{A} = \colim{\ol{S} : \set_0^k, \,\ol{f :
      K\ol{S} \to  A}}{F \ol{S}}
\end{equation}
If $\ol{S} : \set_0^k$ and $\ol{f : K\ol{S} \to A}$, let $j_{\ol{S},
  \ol{f}} : F \ol{S} \to (\mathit{Lan}_KF) \ol{A}$ be the morphism
indexed by $\ol{S}$ and $\ol{f}$ mapping the cocone into the colimit
in Equation~\ref{eq:colim-form}. Then for any $h$-tuple of functions
$\ol{g : A \to B}$, the functorial action $(\mathit{Lan}_K F) \ol{g}$
is the unique function from $(\mathit{Lan}_K F) \ol{A}$ to
$(\mathit{Lan}_K F) \ol{B}$ such that, for all $\ol{S} : \set^k_0$ and
$\ol{f : K\ol{S} \to A}$,\,
\begin{equation}\label{eq:cocone-def}
(\mathit{Lan}_KF) \ol{g} \circ j_{\ol{S},\, \ol{f}} = j_{\ol{S},\,
    \ol{g\, \circ f}}
  \end{equation}
holds.  We can similarly represent left Kan extensions of functors
over $\rel$ in terms of colimits. In that setting we will denote the
morphism mapping the cocone into the colimit $\iota$. We will make
good use of both of these representations in
Section~\ref{sec:term-sem} below.

@@

\[
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\BIC{$\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F} : \Nat^{\Phi,\ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
\DisplayProof
\]
and
\[
\AXC{$\Gamma;\emptyset~|~\Delta \vdash t : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
\DisplayProof
\]

@@

\noindent
is defined to be 
\begin{multline*}
\big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
      \vdash K}\rho[\ol{\alpha := S}]}}\,  
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_S}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}\rho'} \\
\circ\;
  \big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
        \vdash K}\rho[\ol{\alpha := S}]}}\,  
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}f}
\end{multline*}
or, equivalently by naturality,
\begin{multline*}
\big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
      \vdash K}\rho[\ol{\alpha := S}]}}\, 
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}f} \\
\circ\;
  \big(\lambda \ol{S}.\,\textit{Lan}_{\ol{\setsem{\Gamma; \ol\alpha
        \vdash K}\rho[\ol{\alpha := S}]}}\, 
      \setsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_S}]\big)\,
    \ol{\setsem{\Gamma; \Phi \vdash A}\rho}
\end{multline*}

@@

Recalling the notation for the colimit representation of left Kan
extensions from Section~\ref{sec:lke}, we define the set and
relational interpretations of the terms from the introduction and
elimination rules for $\Lan$ types from Section~\ref{sec:ext-calc} by 
\[
\setsem{\Gamma;\emptyset~|~\emptyset \vdash \int_{\ol K,F}
: \Nat^{\Phi, \ol\alpha}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}}\rho\,d
= \eta
\]
and
\[
\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t 
: \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\; G}\rho
= \mu
\]
respectively, where, for each $\ol{N} {\color{red} : ???}$,
$\eta_{\ol{N}}$ is the natural transformation associated with the left
Kan extension
\[\begin{tikzcd} [column sep = large, row sep = large]
\set^{|\ol \alpha|}
\ar[rr, "\lambda \ol{A}.\,\setsem{\Gamma; \Phi, \ol\alpha
    \vdash F}\rho {[\Phi, \ol \alpha := \ol{N}, \ol{A} ]}"{name=F}] 
\ar[dr, "\ol{\lambda \ol{A}.\,\setsem{\Gamma; \ol\alpha \vdash
      K}\rho {[\ol{\alpha := A} ]}}"{left}] && \set \\  
&\set^{|\ol K|} \ar[ur, "\mathit{Lan}_{\ol{\lambda \ol A.\,
      \setsem{\Gamma; \ol\alpha \vdash K}\rho {[\ol{\alpha := A}
  ]}}}\, \lambda \ol A.\, \setsem{\Gamma; \Phi, \ol\alpha \vdash
    F}\rho {[\Phi, \ol \alpha :=  \ol N, \ol A ]}"{right}]
\ar[d, Rightarrow, shorten <= 2mm, shorten >= 1mm, 
          from=F, to=2-2,  "\eta_{\ol{N}}"{}]
\end{tikzcd}\]
and, for all $\ol N {\color{red} : ???}$ and $d {\color{red} : ???}$,
\[\begin{array}{l}
\mu_{\ol{N}\, d} : \lambda \ol{A}.\,
\mathit{Lan}_{\ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}\rho 
  {[\ol{\alpha := A} ]}}} \lambda \ol A.\,\setsem{\Gamma; \Phi,
  \ol\alpha \vdash F}\rho {[\Phi, \ol \alpha := \ol N, \ol A ]}\\[2ex]
\hspace*{0.8in}\to \lambda \ol B.\, \setsem{\Gamma; 
  \Phi, \ol\beta \vdash G} \rho[\Phi, \ol \beta := \ol N, \ol B]
\end{array}\]
is the unique natural transformation such that
\begin{equation}\label{eq:defn-of-mu-sub-N}
(\mu_{\ol{N}\, d} \, \ol{\lambda \ol A.\,\setsem{\Gamma; \ol\alpha \vdash K}
  \rho[\ol{\alpha := A}]}) \circ \eta_{\ol{N}}
= (\setsem{\Gamma; \emptyset \,|\, \Delta \vdash t : \Nat^{\Phi,
  \ol\alpha} \, F \,\, G[\ol{\beta := K}]}\rho \, d)_{\ol{N}}
\end{equation}
as given by the universal property of the left Kan extension.  The
relational interpretations $\relsem{\Gamma;\emptyset~|~\emptyset
  \vdash \int_{\ol K,F} : \Nat^{\Phi, \ol\alpha}\, F\,
  (\Lan^{\ol\alpha}_{\ol{K}}\, F) \ol{K}}$ and
$\setsem{\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F t :
  \Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F) \ol \beta\;
  G}$ are defined entirely analogously.

We conclude this subsection by showing, as promised above, that no
cutting down of set and relational interpretations of $\Lan$-types is
possible. We have:

\begin{prop}\label{prop:full-lan}
Let $F : \set^h \to \set$ and $\ol{K : \set^h \to \set}$ be functors.
If $L : \set^k \to \set$ is a functor such that 
\begin{itemize}
\item $L \ol{A} \subseteq
  (\textit{Lan}_{\ol K} F)\ol{A}$ for all $\ol{A : \set}$,
\item $(\textit{Lan}_{\ol K} F)\ol{f}\,|_{L \ol A}\; x \in L
  \ol{B}$ for all $\ol{f : A \to B}$ and $x \in L \ol A$, and 
\item $\eta_{\ol{A}}\, y \in L ( \ol{K \ol{A}} )$ for all $\ol{A :
  \set}$ and $y \in F \ol{A}$, 
\end{itemize}
then $L = \textit{Lan}_{\ol K} F$.
\end{prop}
\begin{proof}
For all $\ol{A : \set}$ and $z \in (\textit{Lan}_{\ol K} F)\ol{A}$
there exist $\ol{S : \set_0}$, $\ol{f : K \ol{S} \to A}$, and $w \in F
\ol{S}$ such that $z = \iota_{\ol{S}, \ol{f}}\, w$.  Thus, $z =
\iota_{\ol{S}, \ol{f}}\, w = (\textit{Lan}_{\ol K} F) \ol{f}
(\eta_{\ol{S}} \,w)$ by Equation~\ref{eq:cocone-def}. Then by the
third assumption above we have that $\eta_{\ol{S}}\, w \in L (\ol{K
  \ol{S}})$, and by the second assumption above we have that $z \in L
\ol{A}$. This gives $(\textit{Lan}_{\ol K} F)\ol{A} \subseteq L \ol
A$. Finally, by the first assumption above, we therefore have that $L
\ol{A} = (\textit{Lan}_{\ol K} F)\ol{A}$.
\end{proof}
{\color{red} Thus, if $L \subseteq \mathit{Lan}_{\ol K}F$ were the
  interpretation of $\Lan$-types, and if $L$ were to contain the set
  interpretation $\eta$ of ?????, then ????????.

An analogous result holds for relational interpretations of
$\Lan$-types.}

\subsection{Parametricity Results for GADTs}

Having extended our calculus with both $\Lan$-types and terms of such
types, and having given the only sensible set and relational
interpretations for these types and terms, we now need to verify that
these interpretations give rise to a parametric model. The first step
in this process is to extend Lemma~\ref{lem:rel-transf-morph} to
$\Lan$-types by adding a clause for $\Lan$-types to the proof.
Unfortunately, however, Lemma~\ref{lem:rel-transf-morph} does not
extend to arbitrary $\Lan$-types, as the following example shows.

\begin{exa}\label{ex:no-rel-transf}
Consider the type $\emptyset; \alpha \vdash
(\Lan^{\emptyset}_{\onet}\, \onet) \alpha$. The analogue for $\rel$ of
Equation~\ref{eq:colim-form} gives
\[
\begin{array}{rl}
\relsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet} \onet)
  \alpha } \rho [\alpha := (1, 0, 0)] 
&= (\textit{Lan}_{\Eq_1} \Eq_1)(1, 0, 0) \\
&= \colim{f : \Eq_1 \to (1, 0, 0)}{\Eq_1} \\
&= (0, 0, 0)
\end{array}
\]
Here, the last equality holds because there are no morphisms in $\rel$
from $\Eq_1$ to $(1, 0, 0)$, and because $(0,0,0)$ is the initial
object in that category. On the other hand, for the set interpretation
with respect to the first projection of the relation environment
$\rho[\alpha := (1,0,0)]$, Equation~\ref{eq:colim-form} gives
\[
\begin{array}{rl}
\setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet} \onet)
  \alpha } ( \pi_1 ( \rho [\alpha := (1, 0, 0)] ) ) 
&= \setsem{ \emptyset; \alpha \vdash (\Lan^{\emptyset}_{\onet} \onet)
  \alpha } (\pi_1 \rho) [\alpha := 1] \\ 
&= (\textit{Lan}_{1} 1) 1 \\
&= \colim{1 \to 1}{1} \\
&= 1
\end{array}
\]
because $\id_1$ is the unique arrow from $1$ to $1$ in $\set$.
Lemma~\ref{lem:rel-transf-morph} therefore cannot hold.
\end{exa}

The problem in Example~\ref{ex:no-rel-transf} lies in the fact that
the {\color{red} functorial expression} along whose intepretation we
extend contains constants, i.e., {\color{red} subexpressions}
constructed from $\onet$. Indeed, if $\ol{\emptyset; \overline{\alpha}
  \vdash K}$ consists only of polynomial (i.e., sum-of-products)
expressions not containing constants (i.e., formed only from $+$,
$\times$, and the variables in $\alpha$), then
Lemma~\ref{lem:rel-transf-morph} actually does holds for $\Gamma; \Phi
\vdash (\Lan^{\alpha}_{\ol{K}} F)\ol{A}$. This is proved in

\begin{prop}
If $\ol{\emptyset; \overline{\alpha} \vdash K}$ consists only of
polynomial expressions not containing constants, then
\[(\setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
}, \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} },
\relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A} }) \in
[\relenv,RT_0]\]
\end{prop}
\begin{proof}
We need only show that
\begin{equation}\label{eq:former}
\pi_i \,( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
} \rho ) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
  F)\ol{A} } (\pi_i \rho)
\end{equation}
and
\begin{equation}\label{eq:latter}
\pi_i \,( \relsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}} F)\ol{A}
} f ) = \setsem{ \Gamma; \Phi \vdash (\Lan^{\ol \alpha}_{\ol{K}}
  F)\ol{A} } (\pi_i f)
\end{equation}
for all $\rho$, $\rho'$, $f : \rho \to \rho'$, and $i \in \{1, 2\}$.
